---
title: "Data Wrangling"
author: "Wylie Hampson"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm-up:

Look at `us_rent_income` (built-in to the `tidyr` package). Make a tidy version. Then from your tidy version, get it back to the original structure.

```{r}
# Make a wider version:

us_rent_income_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)
```

## Practice some tidyverse functions for data wrangling

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping things together in sequence

### dplyr::filter() to keep or exclude rows based on our conditions

1. Exact match (single condition)

Only keep observations where the island is Biscoe

```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```

2. "and" statement within filter

Only keep observations if the penguin species is Chinstrap and the year is 2008.

```{r}
# All of these methods get the same output.

ex_2 <- penguins %>% filter(species == "Chinstrap", year == 2008)

ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == 2008)

ex_2 <- penguins %>% 
  filter(species == "Chinstrap") %>% 
  filter(year == 2008)
```

3. Exclude Gentoos

Keep observations if the penguin is NOT "Gentoo"

```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Exercise:

Create a subset from `penguins` that only contains observations for Adelie penguins with a flipper length greater than or equal to 185mm

```{r}
exercise_1 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```

4. The %in% operator

Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row.

We want to keep observations where the island is Dream or Torgersen.

```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
```

Neighbor chat: What would actually be happening if in the code, I had used `==` instead of `%in%`.

### select() to select or exlude COLUMNS

5. Keep a single column by name

Only want to keep the species column.

```{r}
# pull() makes a single column from a data frame into a vector

ex_5 <- penguins %>% 
  select(species) %>% 
  pull(species)

# Reminder: rewrite the above without the pipe operator

ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

7. Choose or exclude columns based on position

```{r}
ex_7 <- penguins %>% 
  select(-4)
```

8. A piped sequence

Write a piped sequence where you first: only keep columns for species, body_mass_g, and island, and second: filter to only include observations where body_mass_g is greater than 3600.

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

### mutate() to add or update existing columns, while keeping the ones already there.

9. Add a column that has body_mass_g converted to kg.

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000,
         flipper_length_m = flipper_length_mm / 1000)
```

10. Mutate to update an existing column

```{r}
# Always be careful when overwriting a column. If unsure, just make a new column!

ex_10 <- penguins %>% 
  mutate(island = str_replace(island, patter = "sen", replacement = "HOORAY!"))
```

11. Mutate to update the class of an existing column

Update the class of the bill depth column to character

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm)
```

### case_when(): your favorite if-else

12. Create a new column that contains "small" if penguin flipper length is less than 185mm, "medium" otherwise 

```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: Starting from penguins, create a new column called "bird_bill_bin" that contains: 
- if the bill length is less than 38mm: "short bill"
- if bill length >= 38mm & < 43mm: "medium bill"
- if bill length >= 43: "long bill"

```{r}
exercise_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))
```

### group_by() %>% summarize() to find summary values by group, and return them in a nice summary table

13. group_by() to say there are groups within the species column

```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

ex_13
```

## Practice task

In a single piped sequence:

- Only keep penguins that have a body mass greater than 3500 grams
- Only keep columns for species, flipper_length, and sex
- Create a summary table that contains the mean, sd, max and min value of flipper length by species and sex.

```{r}
exercise_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(flipper_mean = mean(flipper_length_mm),
            flipper_sd = sd(flipper_length_mm),
            flipper_max = max(flipper_length_mm),
            flipper_min = min(flipper_length_mm))

exercise_3
```










